---
title: "Gov 50 Final Project"
author: "Ben Harding"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



Introduction


Although it has long been understood that white men constitute a large portion of suicides in the United States, this project is interested in whether or not those rates of slowed with recent spikes in awareness and sensitivity training. 


Data Visualization 
```{r}
library(ggplot2)
library(dplyr)
suicide_rates <- read.csv("https://data.cdc.gov/api/views/9j2v-jamp/rows.csv?accessType=DOWNLOAD")
 
White_male_data <- suicide_rates |> 
  filter(STUB_LABEL == "Male: White") |>
  filter(AGE == "All ages")

all_persons_data <- suicide_rates |> 
  filter(STUB_LABEL == "All persons") |>
  filter(AGE == "All ages")


White_men_suicide <- ggplot() + 
  geom_line(data = White_male_data, mapping = aes(x = YEAR, y = ESTIMATE, color = "White Male")) + 
  geom_line(data = all_persons_data, mapping = aes(x = YEAR, y = ESTIMATE, color = "All Persons")) + 
  labs(title = "Suicide Rates Over Years: White Males vs. Average", x = "Year", y = "Suicide Rate per 100,000 Resident Population", color = "Population Group") + theme_minimal()
White_men_suicide

```






Plot of Main Analysis and Regression Output  

```{r}
White_male_fit <- lm(ESTIMATE ~ YEAR, data = White_male_data) 

varnames1 <- c("(Intercept)" = "Intercept", 
  "YEAR" = "Suicide Rate Among White Men")
modelsummary::modelsummary(White_male_fit, statistic = "p.value", 
                           gof_map = c("nobs", "r.squared", "adj.r.squared"),
                           coef_map = varnames1)

All_persons_fit <- lm(ESTIMATE ~ YEAR, data = all_persons_data)
varnames1 <- c("(Intercept)" = "Intercept", 
  "YEAR" = "Suicide Rate Among All Persons")
modelsummary::modelsummary(All_persons_fit, statistic = "p.value", 
                           gof_map = c("nobs", "r.squared", "adj.r.squared"),
                           coef_map = varnames1)

white_male_coefficients <- coef(White_male_fit)
all_persons_coefficients <- coef(All_persons_fit)

ggplot() +
  geom_bar(aes(x = "White Males", y = white_male_coefficients[2]), stat = "identity", fill = "blue") +
  geom_bar(aes(x = "All Persons", y = all_persons_coefficients[2]), stat = "identity", fill = "red") +
  labs(title = "Growing Suicide Rates: Disproportionately White Men",
       x = "All Persons vs White Males", y = "Rate of Change in Suicide Rate per Year") +
  theme_minimal()
```
Conclusion 

Rates have not slowed, white men's suicide rates follow similar trends to those of the entire population.

